Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-19 text/sparks.tex
Arrange files into analysis and text folders

\includegraphics[width=2cm]{sparks/6528} &
Shag,
\textit{Phalacrocorax aristotelis},
UK &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused extreme mortality; no longer a nest shortage; population rapidly increased &
\citep{potts1980}\\

\includegraphics[width=2cm]{sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with longterm environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

\includegraphics[width=2cm]{sparks/7115} &
Fur seal,
\textit{Arctocephalus pusillus},
South Africa &
Strong decreases in harvesting, loss of predators, and diamond mining regulations reducing human traffic caused sharp upswings  &
\citep{shaughnessy1982}\\

\includegraphics[width=2cm]{sparks/10113} &
Willow grouse,
\textit{Lagopus lagopus},
UK &
Parasite and predation effects interacted to cause low years  &
\citep{dobson1995}\\

\includegraphics[width=2cm]{sparks/10162} &
Red grouse,
\textit{Lagopus lagopus scoticus},
UK &
Good environmental conditions produced high numbers and vulnerable populations; bad conditions and overcrowding combined to create crashes  &
\citep{mackenzie1952}\\

\includegraphics[width=2cm]{sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was submerged in snow caused population crash &
\citep{newton1998} \\

\includegraphics[width=2cm]{sparks/20579} &
Grey heron,
\textit{Ardea cinerea},
UK &
Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that recovery took three times as long as expected &
\citep{stafford1971} \\

%\includegraphics[width=2cm]{sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
% &
%\citep{brook2006a}\\

%\includegraphics[width=2cm]{sparks/5019} &
%Barbary macaque,  &
% &
%REF\\

%\includegraphics[width=2cm]{sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
% &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
% &
%\citep{stafford1971}\\

>>>>

\includegraphics[width=2cm]{../analysis/sparks/6528} &
Shag,
\textit{Phalacrocorax aristotelis},
UK &
Shortage of nest sites reduced productivity; red-tide event in 1968 caused
extreme mortality; no longer a nest shortage; population rapidly increased &
\citep{potts1980}\\

\includegraphics[width=2cm]{../analysis/sparks/10007} &
Water vole,
\textit{Arvicola terrestris},
UK &
Short-term population cycles from predator interactions combined with longterm
environmental cycle caused sharp downswing  &
\citep{saucy1994}\\

\includegraphics[width=2cm]{../analysis/sparks/7115} &
Fur seal,
\textit{Arctocephalus pusillus},
South Africa &
Strong decreases in harvesting, loss of predators, and diamond mining
regulations reducing human traffic caused sharp upswings  &
\citep{shaughnessy1982}\\

\includegraphics[width=2cm]{../analysis/sparks/10113} &
Willow grouse,
\textit{Lagopus lagopus},
UK &
Parasite and predation effects interacted to cause low years  &
\citep{dobson1995}\\

\includegraphics[width=2cm]{../analysis/sparks/10162} &
Red grouse,
\textit{Lagopus lagopus scoticus},
UK &
Good environmental conditions produced high numbers and vulnerable populations;
bad conditions and overcrowding combined to create crashes  &
\citep{mackenzie1952}\\

\includegraphics[width=2cm]{../analysis/sparks/1235} &
Wren,
\textit{Troglodytes troglodytes},
UK &
Severe winters where food was submerged in snow caused population crash &
\citep{newton1998} \\

\includegraphics[width=2cm]{../analysis/sparks/20579} &
Grey heron,
\textit{Ardea cinerea},
UK &
Severe winters in 1929, 1940--1942, and 1962--1963; 1963 event so severe that
recovery took three times as long as expected &
\citep{stafford1971} \\

%\includegraphics[width=2cm]{../analysis/sparks/20580} &
%Chamois, \textit{Rupicapra rupicapra}, Switzerland &
% &
%\citep{brook2006a}\\

%\includegraphics[width=2cm]{../analysis/sparks/5019} &
%Barbary macaque,  &
% &
%REF\\

%\includegraphics[width=2cm]{../analysis/sparks/9675} &
%Carrot fly (\textit{Psila rosae}, Finland &
% &
%\citep{markkula1965}\\

%\includegraphics[width=2cm]{../analysis/sparks/10139} &
%Grey heron (\textit{Ardea cinerea}), UK &
% &
%\citep{stafford1971}\\

### 2014-10-19 text/supp.tex
Arrange files into analysis and text folders

>>>>

% ------------------------------
% Supplemental Tables
% ------------------------------

\clearpage
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
\input{../analysis/stat-table}
\clearpage

% ------------------------------
% Supplemental Figures
% ------------------------------

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{../analysis/log10-assumed.pdf}
%\caption{Populations in which the abundance was specified as or assumed to be
%recorded as log10 values. The numbers at the front of each panel label are the
%GPDD IDs.}
    %\label{fig:log10-assumed}
%\end{center}
%\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/all-clean-ts-3.pdf}
\caption{
  All filtered times series used in our analysis. The abundances are shown on
  a log10 vertical axis. Colours indicate taxonomic classes. Open black circles
  indicate abundance values that were interpolated. Closed black circles
  indicate single abundance values that were recorded as zero but were set to
  the next lowest observed abundance.
}
    \label{fig:all-ts}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/priors-gomp-base.pdf}
\caption{
  Probability density of the Bayesian priors for the Gompertz models. From left
  to right: (1) per capita growth rate at $\ln$(abundance) = $0$: $\lambda \sim
  \mathrm{Normal}(0, 10^2)$; (2) scale parameter of t-distribution process
  noise: $\sigma \sim \mathrm{Half\mhyphen Cauchy} (0,
  2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2)$; (4) AR1
  correlation coefficient of residuals: $\phi \sim \mathrm{Truncated \mhyphen
    Normal}(0, 1, \mathrm{min.} = -1, \mathrm{max.} = 1)$. Not shown is $b$,
  the density dependence parameter: $b \sim \mathrm{Uniform}(-1, 2)$.
}
  \label{fig:priors}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{../analysis/t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of
  the process error (bottom panels) for a given number of samples (columns)
  drawn from a distribution with a given true $\nu$ value (rows). The red lines
  indicate the true population value. When a small number of samples are drawn
  there may not be samples sufficiently far into the tails to recapture the
  true $\nu$ value; however, heavy tails are still distinguished from normal
  tails in most cases, even with only 25 or 50 samples. TODO switch to nu
  = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz.pdf}
\caption{
  Simulation testing the Gompertz estimation model when the process error draws
  are chosen so that $\nu$ can be estimated close to the true value outside the
  full population model (``effective $\nu$'' within a CV of 0.2 of specified
  $\nu$). The simulation was run across population $\nu$ values (columns) and
  different scenarios (colours): (1) 100 time steps and no observation error,
  (2) 50 time steps and no observation error, (3) 50 time steps and observation
  error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time
  steps with observation error in which the quantity of observation error was
  assumed known. Dashed horizontal lines show the true population values; these
  true values were chosen to represent approximately the median values as
  estimated from the GPDD. Individual dots and lines represent a stochastic
  draws from the true population distribution and a model fitting (median and
  80\% credible intervals).
  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and
  %not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter
  $\sigma$, growth rate parameter $\lambda$, and the density dependence
  parameter $b$.
}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/sim-gompertz-boxplots.pdf}
\caption{
  The distribution of median estimates from the same simulation as described in
  Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The
  whiskers extend to $1.5$ times the interquartile range and outliers are shown
  as dots.
}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/check-sim-box.pdf}
\caption{
  The probability that $\nu < 10$ (i.e.\ the approximate probability of heavy
  tails) for the same simulation scenarios as shown in
  Fig.~\ref{fig:sim-gompertz} and Fig.~\ref{fig:sim-gompertz-boxplots}. Within
  each scenario the dots represent stochastic draws from the true population
  distributions combined with model fits.
}
\label{fig:sim-prob}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{../analysis/ts-gpdd-heavy-eg-log10-base.pdf}
\caption{
  Time series for populations with p($\nu < 10$) $>$ 0.5 using the base
  Gompertz population model. Panels are ordered by increasing p($\nu < 10$).
  Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO
  add dots for upswings and downswings as in Table 1. The labels on each panel
  indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD
  ID number.
}
\label{fig:heavy-ts}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.27\textwidth]{../analysis/order-level-estimates.pdf}
\caption{
  Taxonomic-order-level posterior estimates of probability that $\nu < 10$ from
  a beta regression multilevel model accounting for the effect of time-series
  length. Estimates are at the geometric mean of time series length across all
  the data (approximately 27 time steps). Colour shading refers to the
  taxonomic classes as indicated in Fig.~\ref{fig:nu-coefs} and
  Fig.~\ref{fig:correlates} (yellow for fishes, green for insects, purple for
  birds, and red for mammals). Short vertical line segments indicate median
  posterior estimates. Long dotted vertical line throughout the figure
  indicates the median expected Pr($\nu < 10$) from the prior distribution.
}
\label{fig:order-estimates}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{../analysis/gomp-comparison.pdf}
\caption{
  Estimates of $\nu$ from alternative models plotted against the base Gompertz
  model estimates of $\nu$. Shown are medians of the posterior (dots) and 50\%
  credible intervals (segments). The diagonal line indicates a one-to-one
  relationship. Different colours indicate various taxonomic classes. The
  grey-shaded regions indicate regions of disagreement if $\nu = 10$ is taken
  as a threshold of heavy-tailed dynamics. The Gompertz observation error model
  assumes a fixed standard deviation of observation error of $0.2$ on the log
  scale.
}
\label{fig:alt}
\end{center}
\end{figure}

\clearpage

\begin{figure}[htbp]
\begin{center}
%\includegraphics[width=0.7\textwidth]{../analysis/admb-coefs}
%\includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
\includegraphics[width=0.5\textwidth]{../analysis/stan-beta-correlates}
\caption{
  %Potential correlates of heavy-tailed dynamics.
  %(Top panels) Dots, thick lines, and thin lines represent fixed-effect estimates and their 50\% and 95\% confidence intervals for the covariates listed on the left: $\log \sigma$, $b$, $\lambda$, $\log$(lifespan), $\log$(data set length).
  Main effect posterior densities for potential correlates of heavy-tailed
  dynamics. The response variable was Pr($\nu < 10$) (approximately the
  probability of heavy tails). The beta regression model was fit on a logit
  scale with random intercepts for taxonomic class, taxonomic order, and
  species. All covariates were standardized by subtracting their mean and
  dividing by twice their standard deviation. Therefore, the coefficient values
  represent the expected effect of a change in two standard deviations of the
  raw input variable on the logit-transformed response variable.
}
    \label{fig:correlate-coefs}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.70\textwidth]{../analysis/perc-prob-increase-with-n}
\caption{
Expected percent increase in the probability of detecting heavy tails, Pr($\nu
< 10$), with an additional ten time steps of data. These values are based on
the curve shown in Fig.~\ref{fig:correlates}d, which is based on the
coefficient in Fig.~\ref{fig:correlate-coefs}. For example, we have about
a 12\% increased probability of observing heavy tails in a time series that has
60 time steps vs.\ a time series with 50 time steps.
}
\label{fig:perc-inc-p}
\end{center}
\end{figure}

% \clearpage
%
% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=0.8\textwidth]{../analysis/length-vs-prob-tails}
% \caption{Model predictions. TODO}
%     \label{fig:length-predictions}
% \end{center}
% \end{figure}

\clearpage

\noindent
Example Stan code for a heavy-tailed Gompertz model with AR1 correlated
residuals and a specified level of observation error:

\begin{spacing}{1.15}
\begin{footnotesize}
\begin{verbatim}
data {
  int<lower=3> N;              // number of observations
  vector[N] y;                 // vector to hold ln abundance observations
  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
}
parameters {
  real lambda;                 // Gompertz growth rate parameter
  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
  real<lower=0> sigma_proc;    // process noise scale parameter
  real<lower=2> nu;            // t-distribution degrees of freedom
  real<lower=-1, upper=1> phi; // AR1 parameter
  vector[N] U;                 // unobserved states
  real<lower=0> sigma_obs;     // specified observation error SD
}
transformed parameters {
  vector[N] epsilon;           // error terms
  epsilon[1] <- 0;
  for (i in 2:N) {
    epsilon[i] <- U[i] - (lambda + b * U[i - 1])
                       - (phi * epsilon[i - 1]);
  }
}
model {
  // priors:
  nu ~ exponential(nu_rate);
  lambda ~ normal(0, 10);
  sigma_proc ~ cauchy(0, 2.5);
  phi ~ normal(0, 1);
  // data model:
  for (i in 2:N) {
    U[i] ~ student_t(nu,
                     lambda + b * U[i - 1]
                     + phi * epsilon[i - 1],
                     sigma_proc);
  }
  y ~ normal(U, sigma_obs);
}
\end{verbatim}
\end{footnotesize}

\clearpage
\noindent
Stan code for the multilevel beta regression:
\begin{footnotesize}
\verbatiminput{../analysis/betareg4.stan}
\end{footnotesize}

\clearpage

\noindent
The GPDD IDs used in our analysis:

\begin{footnotesize}
\noindent
{\tt
\input{../analysis/mainids}
}
\end{footnotesize}
\end{spacing}
